### 2022-10-12 AISTATS2023/paper.tex
Review simulation experiments

\subsection{UALE Definition And Interval-Based Formulation}
\label{sec:UALE-definition-1}

@@

%
Intuitively, if we randomly draw a point \(z^*\) from a uniform
distribution \(z^* \sim \mathcal{U}(z_1, z_2)\), the bin-effect
(Eq.~(\ref{eq:mu_bin})) and the bin-uncertainty
(Eq.~(\ref{eq:var_bin})) are the expected average effect and the
expected uncertainty, respectively. Also, if we denote as
\(\mathcal{Z}\) the sequence of \(K+1\) points that partition the
domain of the \(s\)-th feature into \(K\) variable-size intervals,
i.e.  \(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we can reformulate UALE
to:
\begin{equation}
  \label{eq:ALE_2}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \mu}(x_s) = \sum_{k=1}^{k_x} \mu(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}

\begin{equation}
  \label{eq:ALE_accumulated_var}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}
%
where \(k_x\) is the index of the bin such that
\( z_{k_x - 1} \leq x_s < z_{k_x}\). Eq.~(\ref{eq:ALE_2}) and
Eq.~(\ref{eq:ALE_accumulated_var}) are the interval-based formulation
of UALE.

@@

%
which, as shown by\citep{gkolemis22}, is an unbiased estimator of
Eq.~(\ref{eq:mu_bin}) under the assumption that the points are
uniformly distributed inside the intervals. The approximation for the
uncertainty (Eq.~(\ref{eq:var_bin})) from the available instances is
defined as follows:

%
\begin{equation}
  \label{eq:var_bin_approx}
  \hat{\sigma}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|}
\sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left ( \dfdx(\mathbf{x}^i) -
  \hat{\mu}(z_1, z_2) \right )^2
\end{equation}
%
In the Appendix~\ref{sec:proof-1}, we show that
\(\hat{\sigma}^2(z_1, z_2)\) is an unbiased estimator of
\(\sigma^2_*(z_1, z_2) = \frac{\int_{z_1}^{z_2} \mathbb{E}_{\xc|x_s=z}
  \left [ (f^s(z, \xc) - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 -
  z_1} \). In Theorem~\ref{sec:theorem-1} we prove that in the general
case, \(\sigma_*^2(z_1, z_2) \geq \sigma^2(z_1, z_2)\), so
\(\hat{\sigma}^2(z_1, z_2)\) is an overestimation of the bin
uncertainty.

\begin{theorem}
  \label{sec:theorem-1}
  If we define (a) the residual \(\rho(z)\) as the difference between
  the expected effect at \(z\) and the bin-effect, i.e
  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), and (b)
  \(\mathcal{E}(z_1, z_2)\) as the mean squared residual of the bin,
  i.e.
  \(\mathcal{E}(z_1, z_2) = \frac{\int_{z_1}^{z_2}\rho^2(z) \partial
    z}{z_2 - z_1}\), then it holds
\begin{equation}
    \label{eq:bin-uncertainty-proof}
 \sigma_*^2(z_1, z_2) = \sigma^2(z_1, z_2) + \mathcal{E}^2(z_1, z_2)
\end{equation}
  \end{theorem}

@@

The quality of UALE approximation is affected by (a) the population of
samples in each bin and (b) the error term \(\mathcal{E}(z_1, z_2)\)
of each bin. On the one hand, we favor a partitioning of the feature
domain in large-bins to allow a robust estimation of
\(\hat{\mu}(z_1, z_2), \hat{\sigma}(z_1, z_2)\) from a sufficient
population of samples. On the other hand, we want to minimize the
cummulative bin-error, i.e.,
\( \mathcal{E}^2_{\mathcal{Z}} = \sum_{k=1}^K\mathcal{E}^2(z_1, z_2)
\Delta z_k\), where \(\mathcal{Z} = \{z_0, \cdots, z_K\}\) and
\(\Delta z_k = z_k - z_{k-1}\). For convenience, we refer to
\(\mathcal{E}^2_{\mathcal{Z}}\) as accumulated error. We search for a
partitioning \(\mathcal{Z}_*\) that balances this trade-off.

Due to the Corollary~\ref{sec:coroll}, minimizing
\( \mathcal{E}^2_{\mathcal{Z}} \) is equivalent to minimizing
\(\sigma_*^2(z_{k-1}, z_k)\Delta z_k\), which can be directly
estimated by \(\hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \).

\begin{corollary}
  \label{sec:coroll}
  A bin-splitting \(\mathcal{Z}\) minimizes the accumulated if and
  only if it also minimizes the accumulated error.
  \begin{equation}
  \begin{gathered}
    \label{eq:2}
    \argmin_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \mathcal{E}^2_{\mathcal{Z}}  = \argmin_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \sum_{k=1}^K\sigma_*^2(z_1, z_2) \Delta z_k 
  \end{gathered}
  \end{equation}
\end{corollary}

\begin{proof}
~\label{sec:coroll-1}
  The proof is at Appendix~\ref{sec:proof-of-coroll}
  \end{proof}

@@

\begin{equation}
  \label{eq:opt}
\begin{aligned}
  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \tau_k \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
  \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
  & \tau_k = 1 - \alpha \frac{|S_k|}{N} \\
  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
                                     & z_0 = x_{s,min}\\
                                     & z_K = x_{s, max}
\end{aligned}
\end{equation}
%
The objective \(\mathcal{L}\) searches for a partitioning
\(\mathcal{Z}_* = \{ z_0, \ldots, z_K \} \) with low aggregated error
\(\mathcal{E}_{\mathcal{Z}}^2\). In case of many partitionings with
similar aggregate-error, the term \(\tau_K\) favors the partitioning
with more points per bin (with wider bins). The constraint of at least
\(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
\textit{robust} estimation. The user can choose to what extent they
favor the creation of wide bins through (a) the parameter \(\alpha\)
that controls the discount \(\tau_k\) and (b) the parameter
\(N_{\mathtt{PPB}}\) that sets the minimum population per bin. For
providing a rough idea, in our experiments we set \(\alpha = 0.2\)
which means that the discount ranges between \([0\%, 20\%]\) and
\(N_{\mathtt{PPB}} = \frac{N}{20}\), where \(N\) is the dataset size.

For solving the optimization problem of~Eq.\ref{eq:opt} we add some
constraints. First, we set a threshold \(K_{max}\) on the maximum
number of bins which, in turn, defines the minimum bin-width,
i.e. \(\Delta x_{min} = \frac{x_{s, max} - x_{s,min}}{K_{max}}\).
Based on that, we restrict the bin-limits to be multiples of the
minimum width, i.e.
\(z_k = k\cdot \Delta x_{min}, \text{where } k \in \{0 , \cdots,
K_{max}\} \). In this discretized solution space, we find the global
optimal using dynamic programming.

\subsubsection{Solve Bin-Splitting with Dynamic Programming}
\label{sec:dynamic-programing}

@@

In XAI it is common~\citep{aas2021explaining, herbinger2022repid} to
use simulation examples for compering methods, because the
data-generating distribution \(p(\mathbf{X})\) and the predictive
function \(f(\cdot)\) are known, and, therefore, the ground-truth is
accesible. We follow this common XAI practice and we split the
evaluation in two groups of examples. The first group
(Section~\ref{sec:simulation-examples-1}) compares UALE definition
against PDP-ICE in modeling the average effect and the
uncertainty. The second group
(Section~\ref{sec:simulation-examples-2}) compares UALE approximation
using the automatic bin-splitting against the fixed-size alternative,
in estimating the average effect and the uncertainty.

@@

In this simulation, we compare UALE against PDP-ICE. The aim of this
example is to highlight that when features are correlated, PDP and ICE
fail in quantifying the average effect and the heterogeneity, even
when the black-box function is relatatively simple.

@@

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $f_{\mathtt{lin}} < 0.5$ }\\
                    0.5 - f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $0.5 \leq f_{\mathtt{lin}} < 1$}\\
                    \alpha f_{\mathtt{int}} &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_{\mathtt{int}}(\mathbf{x}) = x_1x_3\). The linear term
\(f_{\mathtt{lin}}\) includes the two correlated features and the
interaction term \(f_{\mathtt{int}}\) the two non-correlated.  We
evaluate the effect computed by UALE and PDP-ICE in three cases; (a)
without interaction (\(\alpha=0\)) and equal weights (\(a_1=a_2\)),
(b) without interaction (\(\alpha=0\)) and different weigths
(\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha \neq 0\)) with
equal weights (\(a_1=a_2\)). In all cases, we provide the ground-truth
average effect and uncertainty computed with analytical derivations
(proofs at the Appendix) and we compare it against the approximations
provided by each method, after generating \(N=500\) samples.

@@

In this set-up, \(\alpha=0\) (no interaction) and the \(a_1=a_2=1\)
(equal weights). The ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\)
is: \(x_1\) in \([0, 0.25]\), \(-x_1\) in \([0.25, 0.5]\) and \(0\) in
\([0.5, 1]\). The uncertainty is
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) in \([0,1]\) because \(x_1\)
does not interact with ohter variables. In
Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP effect is
wrong and ICE plots imply the existence of heterogeneous effects. In
contrast, UALE models correctly both the average effect and the zero
uncertainty and approximates it accurately finding a zero aggregate
residual error partitioning of \(x_1\) domain.

@@

In this set-up, \(\alpha=0\) (no interaction), \(a_1 = 2\) and
\(a_2= 0.5\) (non-equal weights). The non-equal weights have
implications at both the gradient and the interval of the piece-wise
linear regions, i.e., \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(2x_1\) in
\([0, 0.2]\), \(-2x_1\) in \([0.2, 0.4]\) and \(0\) in \([0.4,
1]\). As before, the ground-truth uncertainty is
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) because \(x_1\) does not
interact with other features.  In Figure~\ref{fig:ex-synth-1-2}, we
observe that PDP estimation is completely opposite to the ground-truth
effect, i.e.~negative in the region \([0, 0.2)\), positive in
\([0.2, 0.4)\), and the ICE erroneously implies the existence of
heterogeneous effects. As before, ALE quantifies correctly the ground
truth effect, the zero-uncertainty and partioning correctly the
\(x_1\) domain.

@@

\paragraph{Uncertainty, Equal weights.}

In this set-up, we activate the interaction term, i.e.  \(a=1\),
keeping the weights equal \(a_1=a_2=1\). The interaction term provokes
heterogeneous effects for features \(x_1\) and \(x_3\), because the
local effects at \(x_1\) depend on the unknown value of \(X_3\) and
vice-versa. The effect of \(x_2\) has zero-uncertainty since it does
not appear in any interaction term.  As we observe in
Figure~\ref{fig:ex-synth-1-3}, UALE correcly models the average effect
and the uncertainty in all cases.  On the other hand, PDP-ICE
quantifies correctly the average effect and the uncertainty only in
the case of feature \(x_3\), because \(X_3\) is independent from other
features. For the correlated features (\(x_1, x_2\)) both the average
effect (PDP) and the uncertainty (ICE) are wrong.

@@

Despite the model's simplicity, PDP-ICE fail in modeling both the
average effect and the uncertainty. The error is due to the method's
formulation, i.e.~the use of the marginal distribution that ignores
correlations between features, and not due to poor approximation
because of limited samples. In contrast, UALE models both correctly
and estimates them accurately extracting the regions with constant
effect.  The examples above do not cover the case of an interaction
term between correlated features, for example a term \(x_1x_2\),
because there is an open debate about the ground-truth effect in this
case~\citep{Gromping2020MAEP}.

@@

In this simulation, we illustrate the advantages of automatic
bin-splitting against the fixed-size alternative. For this reason, we
generate a very big dataset applying dense sampling (\(N=10^6\)) and
we treat the estimation with dense fixed-size bins (\(K=10^3\)) as the
ground-truth UALE. Afterwards, we generate less samples (\(N=500\))
and we compare the fixed-size estimation (for many different \(K\))
against UALE automatic bin-splitting. In all set-ups, we sample from
\(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\) and
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We denote as
\(\mathcal{Z^{\mathtt{DP}}} = \{z^{\mathtt{DP}}_{k-1}, \cdots,
z^{\mathtt{DP}}_{K}\}\) the sequence obtained by automatic
bin-splitting and with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size
splitting with \(K\) bins. The evaluation metrics we report are the
average number of \(t = 30\) independent runs, using each time
\(N=500\) different samples.

@@

The evaluation is done with regard to two metrics counting the mean
error per bin, where the bin error is definde as the absolute
difference of the approximation from the ground-truth. The first
metric, Eq.~\eqref{eq:eval_met_1}, counts the mean bin error wrt
average effect and the second, Eq.~\eqref{eq:eval_met_2} the mean bin
error wrt uncertainty:

\begin{equation}
  \label{eq:eval_met_1}
  \mathcal{L}_{\mathtt{DP|K}}^{\mu} = \frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} | \mu(z_k- z_{k-1}) - \hat{\mu}(z_k-
  z_{k-1}) |
\end{equation}

\begin{equation}
  \label{eq:eval_met_2}
  \mathcal{L}_{\mathtt{DP|K}}^{\sigma} = \frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
    \mathcal{Z}^{\mathtt{DP|K}}} | \sigma(z_k- z_{k-1}) - \hat{\sigma}(z_k-
  z_{k-1}) |
\end{equation}

The ground truth \(\mu_k\) (\(\sigma_k\)) is the mean value of the
bin-effects (bin-uncertainties) of the bins inside the interval
defined by the wide bin. For better interpretation of the uncertainty
error, we also provide the mean (per bin) error term
\(\mathcal{L}^{\rho}_{\mathtt{DP|K}} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in \mathcal{Z}^{\mathtt{DP|K}}} \mathcal{E}(z_{k-1}, z_k) \).

@@

In this set-up, \(f(\mathbf{x}) = a_1x_1 + x_1x_2\) is a
piecewise-linear function with \(5\) different-width regions, i.e.,
\(a_1\) equals to \(=\{2, -2, 5, -10, 0.5\}\) in the intervals defined
by the sequence \(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). As we observe, in
Figure~\ref{fig:ex-synth-2-1}, UALE's approximation is better than all
fixed-size alternatives, in both mean effect (\(\mathcal{L}^{\mu}\))
and uncertainty (\(\mathcal{L}^{\sigma}\)). For understanding the
importance of variable-size splitting, we analyze the fixed-bin
error. In the top-right of Figure~\ref{fig:ex-synth-2-1}, we observe
the Law of Large Numbers (LLN); \(\mathcal{L}^{\mu}_{\mathtt{K}}\) and
\(K\) have positive correlation because as \(\Delta x\) becomes
smaller less points contributing to each estimation. The
interpretation of \(\mathcal{L}^{\sigma}\) is more complex. Given that
the smallest piecewise-linear region is \(\Delta x = 0.05 \) any
\(K\) that is not a multiple of \(\frac{1}{\Delta x} = 20\), adds
nuisance uncertainty \(\mathcal{L}^{\rho}_{\mathtt{K}} > 0\) leading
to a biased uncertainty approximation. For these \(K\), uncertainy
error decreases for bigger \(K\) because the length of the bins that
lie in the limit between two piecewise linear regions becomes
smaller. For \(K\) that are multiple of \(20\), i.e.
\(K=\{20, 40, 60, 80, 100\}\), the uncertainty approximation is
unbiased \(\mathcal{L}^{\rho}_{\mathtt{K}} \approx 0\). Here, as in
\(\mathcal{L}^{\mu}\), uncertainty estimation is worse as \(K\) grows
larger (LLN). As we observe in the top left of
Figure~\ref{fig:ex-synth-2-1}, automatic bin-splitting separates the
fine-grain bins, e.g.~regions \([0.4, 0.45]\), \([0.45, 0.5]\), and
unites constant-effect regions into a single bin, e.g.~region
\([0.5, 1]\).

>>>>

\subsection{UALE Definition and Interval-Based Formulation}
\label{sec:UALE-definition-1}

@@

%
Intuitively, if we randomly draw a point \(z^*\) from a uniform
distribution \(z^* \sim \mathcal{U}(z_1, z_2)\), the bin-effect
(Eq.~(\ref{eq:mu_bin})) and the bin-uncertainty
(Eq.~(\ref{eq:var_bin})) are the expected average effect and the
expected uncertainty, respectively. Also, if we denote as
\(\mathcal{Z}\) the sequence of \(K+1\) points that partition the
domain of the \(s\)-th feature into \(K\) variable-size intervals,
i.e.  \(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we provide the
interval-based formulation of UALE:

\begin{equation}
  \label{eq:ALE_2}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \mu}(x_s) = \sum_{k=1}^{k_x} \mu(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}

\begin{equation}
  \label{eq:ALE_accumulated_var}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}
%
where \(k_x\) is the index of the bin such that
\( z_{k_x - 1} \leq x_s < z_{k_x}\). Eq.~(\ref{eq:ALE_2}) and
Eq.~(\ref{eq:ALE_accumulated_var}) are piecewise-linear approximation
of Eq.~(\ref{eq:ALE}) and Eq.~(\ref{eq:ALE_acc_unc}).

@@

%
which is an unbiased estimator of Eq.~(\ref{eq:mu_bin}) under the
assumption that the points are uniformly distributed inside the
interval. The approximation for the uncertainty
(Eq.~(\ref{eq:var_bin})) from the available instances is defined as
follows:

%
\begin{equation}
  \label{eq:var_bin_approx}
  \hat{\sigma}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|}
\sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left ( \dfdx(\mathbf{x}^i) -
  \hat{\mu}(z_1, z_2) \right )^2
\end{equation}
%
In the Appendix~\ref{sec:proof-1}, we show that
\(\hat{\sigma}^2(z_1, z_2)\) is an unbiased estimator of
\(\sigma^2_*(z_1, z_2) = \frac{\int_{z_1}^{z_2} \mathbb{E}_{X_c|X_s=z}
  \left [ (f^s(z, X_c) - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 -
  z_1} \). In Theorem~\ref{sec:theorem-1} we prove that in the general
case, \(\sigma_*^2(z_1, z_2) \geq \sigma^2(z_1, z_2)\), so
\(\hat{\sigma}^2(z_1, z_2)\) is an overestimation of the bin
uncertainty.

\begin{theorem}
  \label{sec:theorem-1}
  If we define (a) the residual \(\rho(z)\) as the difference between
  the expected effect at \(z\) and the bin-effect, i.e
  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\) and (b)
  \(\mathcal{E}(z_1, z_2)\) as the mean squared residual of the bin,
  i.e.
  \(\mathcal{E}(z_1, z_2) = \frac{\int_{z_1}^{z_2}\rho^2(z) \partial
    z}{z_2 - z_1}\), then it holds
\begin{equation}
    \label{eq:bin-uncertainty-proof}
 \sigma_*^2(z_1, z_2) = \sigma^2(z_1, z_2) + \mathcal{E}^2(z_1, z_2)
\end{equation}
  \end{theorem}

@@

The quality of UALE approximation is affected by (a) the population of
samples in each bin and (b) the error term \(\mathcal{E}(z_1, z_2)\)
of each bin. On the one hand, we favor a partitioning of the feature
domain in large-bins to allow a robust estimation of
\(\hat{\mu}(z_1, z_2), \hat{\sigma}(z_1, z_2)\) from a sufficient
population of samples. On the other hand, we want to minimize the
cumulative bin-error, i.e.,
\( \mathcal{E}^2_{\mathcal{Z}} = \sum_{k=1}^K\mathcal{E}^2(z_1, z_2)
\Delta z_k\), where \(\mathcal{Z} = \{z_0, \cdots, z_K\}\) and
\(\Delta z_k = z_k - z_{k-1}\). We search for a partitioning that
balances this trade-off.

Corollary~\ref{sec:coroll} shows that minimizing
\( \mathcal{E}^2_{\mathcal{Z}} \) is equivalent to minimizing
\(\sum_{k=1}^K \sigma_*^2(z_{k-1}, z_k)\Delta z_k\), which can be directly
estimated from \(\sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \).

\begin{corollary}
  \label{sec:coroll}
  A bin-splitting \(\mathcal{Z}\) that minimizes the accumulated error
  if and only if it also minimizes
  \(\sum_{k=1}^K\sigma_*^2(z_1, z_2) \Delta z_k \)
\end{corollary}

\begin{proof}
  ~\label{sec:coroll-1} The proof is based on the observation that
  \(\sum_{k=1}^K \sigma^2(z_{k-1}, z_k) \Delta z_k = \sigma^2(z_0,
  z_K) (z_K - z_0)\) which is independent of the bin-splitting. A more
  detailed proof is provided in the Appendix~\ref{sec:proof-of-coroll}.
 \end{proof}

@@

\begin{equation}
  \label{eq:opt}
\begin{aligned}
  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \tau_k \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
  \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
  & \tau_k = 1 - \alpha \frac{|S_k|}{N} \\
  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
                                     & z_0 = x_{s,min}\\
                                     & z_K = x_{s, max}
\end{aligned}
\end{equation}
%
The objective \(\mathcal{L}\) searches for a partitioning
\(\mathcal{Z}_* = \{ z_0, \ldots, z_K \} \) with low aggregated error
\(\mathcal{E}_{\mathcal{Z}}^2\). In case of many partitionings with
similar aggregate-error, the term \(\tau_K\) favors the partitioning
with more points per bin, i.e., with wider bins. The constraint of at
least \(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
\textit{robust} estimation. The user can choose to what extent they
favor the creation of wide bins through (a) the parameter \(\alpha\)
that controls the discount \(\tau_k\) and (b) the parameter
\(N_{\mathtt{PPB}}\) that sets the minimum population per bin. For
providing a rough idea, in our experiments we set \(\alpha = 0.2\)
which means that the discount ranges between \([0\%, 20\%]\) and
\(N_{\mathtt{PPB}} = \frac{N}{20}\), where \(N\) is the dataset size.

For solving the optimization problem of~Eq.\ref{eq:opt} we add some
constraints. First, we set a threshold \(K_{max}\) on the maximum
number of bins which, in turn, defines the minimum bin-width,
i.e. \(\Delta x_{min} = \frac{x_{s, max} - x_{s,min}}{K_{max}}\).
Based on that, we restrict the bin-limits to the multiples of the
minimum width, i.e.
\(z_k = k\cdot \Delta x_{min}, \text{where } k \in \{0 , \cdots,
K_{max}\} \). In this discretized solution space, we find the global
optimum using dynamic programming. \todo{Add description} See also
Appendix~\ref{sec:dynamic-programming}.

% \begin{algorithm}
%   \end{algorithm}

@@

Similarly to~\citep{aas2021explaining, herbinger2022repid} we first
use simulation examples for evaluation. The first group
(Section~\ref{sec:simulation-examples-1}) compares UALE against
PDP-ICE in modeling the average effect and the uncertainty. The second
group (Section~\ref{sec:simulation-examples-2}) compares UALE
approximation with optimal bin-splitting against the fixed-size
alternative.

@@

The aim of this example is to highlight that when features are
correlated, PDP and ICE fail in quantifying the average effect and the
heterogeneity, even when the black-box function is relatatively
simple.

@@

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $f_1(\mathbf{x}) < \frac{1}{2}$ }\\
                    \frac{1}{2} - f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $\frac{1}{2} \leq f_2(\mathbf{x}) < 1$}\\
                    \alpha f_1(\mathbf{x}) &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_1(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_2(\mathbf{x}) = x_1x_3\). The linear term \(f_1\) includes the two
correlated features and the interaction term \(f_2\) the two
non-correlated.  We evaluate the effect computed by UALE and PDP-ICE
in three cases; (a) without interaction (\(\alpha=0\)) and equal
weights (\(a_1=a_2\)), (b) without interaction (\(\alpha=0\)) and
different weigths (\( a_1 \neq a_2 \)) and (c) with interaction
(\(\alpha \neq 0\)) with equal weights (\(a_1=a_2\)).

In all cases, we assume that because the \todo{give ground truth
  effect}. We compare it against the approximations provided by each
method, after generating \(N=500\) samples.

@@

In this set-up, \(\alpha=0\) (no interaction) and the \(a_1=a_2=1\)
(equal weights). The ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\)
is: \(x_1\) in \([0, 0.25]\), \(-x_1\) in \([0.25, 0.5]\) and \(0\) in
\([0.5, 1]\). The uncertainty is
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) in \([0,1]\) because \(x_1\)
does not interact with ohter variables. In
Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP effect is
wrong and ICE plots show heterogeneous effects. In contrast, UALE
models correctly both the average effect and the
uncertainty. Furthermore, we observe that bin-splitting is accurate
(zero cumulative error \(\mathcal{E}_{\mathcal{Z}}\)).

@@

In this set-up, \(\alpha=0\) (no interaction), \(a_1 = 2\) and
\(a_2= 0.5\) (non-equal weights). The non-equal weights have
implications at both the gradient and the interval of the piece-wise
linear regions, i.e., \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(2x_1\) in
\([0, 0.2]\), \(-2x_1\) in \([0.2, 0.4]\) and \(0\) in \([0.4,
1]\). As before, the ground-truth uncertainty is
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) because \(x_1\) does not
interact with other features.  In Figure~\ref{fig:ex-synth-1-2}, we
observe that PDP estimation is opposite to the ground-truth effect,
i.e.~negative in the region \([0, 0.2)\), positive in \([0.2, 0.4)\),
and the ICE erroneously implies the existence of heterogeneous
effects. As before, ALE quantifies correctly the ground truth effect,
the zero-uncertainty and partioning correctly the \(x_1\) domain.

@@

\paragraph{With Interaction, Equal weights.}

In this set-up, we activate the interaction term, i.e.  \(a=1\),
keeping the weights equal \(a_1=a_2=1\). The interaction term provokes
heterogeneous effects for features \(x_1\) and \(x_3\), because the
local effects at \(x_1\) depend on the unknown value of \(x_3\) and
vice-versa. The effect of \(x_2\) has zero-uncertainty since it does
not appear in any interaction term.  As we observe in
Figure~\ref{fig:ex-synth-1-3}, UALE correcly models the average effect
and the uncertainty in all cases.  On the other hand, PDP-ICE
quantifies correctly the average effect and the uncertainty only in
the case of feature \(x_3\), because it is independent from other
features. For the correlated features \(x_1\) and \(x_2\) both the
average effect (PDP) and the uncertainty (ICE) are wrong.

@@

Despite the model's simplicity, PDP-ICE fail in modeling both the
average effect and the uncertainty. The error is due to the method's
formulation, i.e.~the use of the marginal distribution that ignores
correlations between features, and not due to poor approximation
because of limited samples. In contrast, UALE models both correctly
and estimates them accurately using an optimal bin splitting in terms
of Eq.~(\ref{eq:opt}). The examples above do not cover the case of an interaction term
between correlated features, for example a term \(x_1x_2\), because
there is an open debate about the ground-truth effect in this
case~\citep{Gromping2020MAEP}.

@@

In this simulation, we illustrate the advantages of automatic
bin-splitting against the fixed-size alternative. For this reason, we
generate a very big dataset applying dense sampling (\(N=10^6\)) and
we treat the estimation with dense fixed-size bins (\(K=10^3\)) as the
ground-truth UALE. Afterwards, we generate fewer samples (\(N=500\))
and we compare the fixed-size estimation (for several \(K\)) against
UALE automatic bin-splitting. In all set-ups, we sample from
\(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\) and
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We denote as
\(\mathcal{Z^*} = \{z^*_0, \cdots,
z^*_K\}\) the sequence obtained by automatic
bin-splitting and with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size
splitting with \(K\) bins. The evaluation metrics we report are the
average number of independent runs \(t = 30\), using each time
\(N=500\) different samples.

@@

Evaluation is perform in terms of the Mean Absolute Error (MAE) of the
estimation of \(\mu\) and \(\sigma\) accross bins, i.e.,

% The evaluation is done with regard to two metrics counting the mean
% error per bin, where the bin error is definde as the absolute
% difference of the approximation from the ground-truth. The first
% metric, Eq.~\eqref{eq:eval_met_1}, counts the mean bin error wrt
% average effect and the second, Eq.~\eqref{eq:eval_met_2} the mean bin
% error wrt uncertainty:

\begin{equation}
  \label{eq:eval_met_1}
  \mathcal{L}^{\mu} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
  \mathcal{Z}} | \mu(z_{k-1}, z_k) - \hat{\mu}(z_{k-1}, z_k) |
\end{equation}

\begin{equation}
  \label{eq:eval_met_2}
  \mathcal{L}^{\sigma} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
    \mathcal{Z}} | \sigma(z_{k-1}, z_k) - \hat{\sigma}(z_{k-1}, z_k) |
\end{equation}

The ground truth UALE is the average of the dense fixed-size bins that
are within in the interval \([z_{k-1}, z_k]\). For better
interpretation of \(\mathcal{L}^{\sigma}\), we also provide the mean
error term
\(\mathcal{L}^{\rho} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
  \mathcal{Z}} \mathcal{E}(z_{k-1}, z_k) \).

@@

In this set-up, \(f(\mathbf{x}) = a_1x_1 + x_1x_2\) is a
piecewise-linear function with \(5\) different-width regions, i.e.,
\(a_1\) equals to \(\{2, -2, 5, -10, 0.5\}\) in the intervals defined
by the sequence \(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\).

As we observe in the top left of Figure~\ref{fig:ex-synth-2-1}, UALE's
bin-splitting separates the fine-grained bins, e.g.~regions
\([0.4, 0.45]\), \([0.45, 0.5]\), and unites (most) constant-effect
regions into a single bin, e.g.~region \([0.5, 1]\). This explains the
fact that UALE achieves lower MAE \(\mathcal{L}^{\mu}\),
\(\mathcal{L}^{\sigma}\) compared to fixed-size bins for all \(K\).

% We observe that in
% this case all intervals are multiples of \(0.05\). As we observe, in
% Figure~\ref{fig:ex-synth-2-1}, UALE's approximation is better than all
% fixed-size alternatives (for all \(K\)), in both mean effect
% \(\mathcal{L}^{\mu}\) and uncertainty \(\mathcal{L}^{\sigma}\). For
% understanding the importance of variable-size splitting, we analyze
% the fixed-bin error. In the top-right of
% Figure~\ref{fig:ex-synth-2-1}, we observe a positive trend between
% \(\mathcal{L}^{\mu}\) and \(K\), because as \(K\) becomes larger fewer
% points contribute to the bin estimations, leading to a worse
% approximation. The interpretation of \(\mathcal{L}^{\sigma}\) is more
% complex. For \(K\) that are multiples of \(20\), i.e.
% \(K=\{20, 40, 60, 80, 100\}\), the mean error term
% \(\mathcal{L}^{\rho}\) is low as we see in the bottom-right of
% Figure~\ref{fig:ex-synth-2-2}. In these cases, we also observe in the
% bottom-left figure, that \(L\) is increasing with \(K\) due to fewer
% points in each bin. For all other values of \(K\), the \(L\) is mainly
% affected by the mean error term \(\mathcal{E}\).
